---
title: "Piping Practice"
author: "Shannon Ellis"
date: "9/11/2018"
output: html_document
---

The goal of this document is to help get more comfortable tidying data using the functions available in the `dplyr` package.

### `dplyr` functions

Before we get started, let's review the main `dplyr` functions you'll need to use throughout this lesson:

* glimpse() - get an overview of what’s included in dataset
* filter() - filter rows
* select() - select, rename, and reorder columns
* arrange() - reorder rows
* mutate() - create a new column
* group_by() - group variables
* summarize() - summarize information within a dataset
* left_join() - combining data across data frame

### pipe review

Also, we want to take a minute to highlight how you tie all these functions together: the pipe operator: `%>%`. This operator takes whatever the *output* of the function before it and uses that as the *input* for the next function..

#### A good pipe 

Here, we'll use something that *looks* like code but isn't actually coe as an example:

```
shoe %>% 
  put_on() %>%
  tie_it()
```

Here, I'm saying, start with a shoe *then* (`%>%`) put it on your foot using hte `put_on` function. Then, take that shoe that's now on your foot (the output from shoe %>% `put_on()`) and tie it (`tie_it()`).

#### A broken pipe 

Using this same example, what if I instead said:

```
shoe %>%
  put_on()
tie_it()
```

Here, we've removed the pipe between `put_on()` and `tie_it()`.

Well, this says, put the shoe on *then* tie that shoe (because there's a pipe there)

*But* the `tie_it()` line doesn't have an input. The `tie_it` function doesn't know what to tie...

#### Two separate pipes

Okay, one final toy example. What if we said...

```
shoe %>%
  put_on()

shoe %>%
  tie_it()
```

Well, here, the first two lines says take the shoe and put it on. But the next line says take the shoe (that's not on the foot yet) and tie it.

This would get you a shoe on a foot *and* a shoe that was not on a foot but that was tied. Not exactly what we were looking for.

#### Back to the good pipe 

In order to get a shoe, on a foot, that is tied, you need a pipe to take the output from the step prior and "pipe it" into the next function:

```
shoe %>%
  put_on() %>%
  tie_it()
```

Now that we've reviewed the functions in `dplyr` *and* what a pipe is, let's get down to writing some code!

### The Datasets

In these exercies we'll be using two datasets: `mtcars` and `msleep`.

`msleep` is available from the `ggplot2` package. 


```{r load-package}
## Install and load the ggplot2 package
install.packages("ggplot2")
install.packages("dplyr")
library(ggplot2)
library(dplyr)
```


## Looking at the Datasets

Let's get an idea of what variables are available in each dataset. Use the functions `glimpse()` and `View()` to get a sense of both the `mtcars` and the `msleep` dataset. 

**Think about the differences between the `glimpse()` and `View()` functions**

```{r view-data}
## Get acquainted with the dataset
glimpse(mtcars)
glimpse(msleep)
View(mtcars)
View(msleep)
```

## Filtering Rows

Now that you have some idea of what variables and how many observations are included in each dataset, we'll get down to using more `dplyr` functions!

Sometimes you have a dataset but are only interested in working with some *part* of the dataset. To do this, you want to filter the full dataset to only include the rows you're interested in. 

For example, the `msleep` dataset includes data about lots of different mammals. what if you only were interested in studying carnivores?

```{r filtering-rows-msleep}
## Filter the `msleep` dataset to only include carnivorous animals
## call this new dataframe `carnivores`
carnivores <- filter(msleep, order == "carnivora")
```

```{r observations-carnivores}
## How many observations are in this new `carnivores` dataframe?
0
```


Or, what about in the `mtcars` dataset. What if you weren't interested in working with any 4-cylinder cars? 

```{r filtering-rows-mtcars}
## Filter the `mtcars` dataset to remove all 4 cylinder cars
## call this new dataframe `big_cars`
big_cars <- filter(mtcars, cyl == "4")
```

```{r observations-big-cars}
## How many observations are in this new `big_cars` dataframe?
11
```

## Reordering rows

You should now have two filtered datasets: `carnivores` and `big_cars`. 

In addition to filtering out rows, you often want your data in a different order than they are currently.

For the `carnivores` dataframe, it's reasonable that if you're interested in how long each animal sleeps, you'd want to sort the rows by this variable (`msleep`). 

```{r reordering-rows-carnivores}
## Reorder the rows of the `carnivores` dataset 
## Have the shortest sleepers first and the longest sleepers last
## again assign this dataframe to the variable `carnivores`
carnivores <- arrange(carnivores, sleep_total)
```

Let's do something similar with our new `big_cars` dataframe.

Here we want to order our rows by how heavy the cars are.

```{r reordering-rows-big-cars}
## Reorder the rows of the `big_cars` dataset 
## Have the largest cars first and the smallest cars last
## again assign this dataframe to the variable `big_cars`
big_cars <- arrange(big_cars, wt)
```

### Piping

So far we've filtered our original datasets (`mtcars` and `msleep`) and re-ordered the rows. We've done that in two different steps. However, this can all be done at once, using pipes.

Recreate the `carnivores` dataframe you created above (filtering and re-ordering), but do it all at once using pipes

```{r pipe-carnivores}
## Recreate the `carnivores` object
## Do it all at once
## using pipes
msleep %>%
  filter(order == "carnivora") %>%
  arrange(carnivores, sleep_total)
```

Now, recreate the `big_cars` dataframe you created above (filtering and re-ordering), but do it all at once using pipes

```{r pipe-big-cars}
## Recreate the `big_cars` object
## Do it all at once
## using pipes
mtcars %>%
  filter(cyl == "4") %>%
  arrange(wt)
```
